R Self Assessment
=================

[Questions](http://www.biostat.jhsph.edu/~rpeng/coursera/selfquiz/quiz.html) | [Answers](http://www.biostat.jhsph.edu/~rpeng/coursera/selfquiz/quiz_answer.html)

1. Download and install R from the Comprehensive R Archive Network. Make sure to choose a version that is appropriate for your computing platform (Windows, Mac, or Unix/Linux)

2. Download the dataset available located on this web page and load it into R with the read.csv function. Assign the output of read.csv to an object named dataset.
```{r}
dataset <- read.csv("~/Desktop/selfquiz-data.csv")
```

3. What are the column names of the data frame?
```{r}
colnames(dataset)
```

4. What are the row names of the data frame?
```{r}
rownames(dataset)
```

5. Extract the first 6 rows of the data frame and print them to the console
```{r}
head(dataset)
```

6. How many observations (i.e. rows) are in this data frame?
```{r}
nrow(dataset)
```

7. Extract the last 6 rows of the data frame and print them to the console
```{r}
tail(dataset)
```

8. How many missing values are in the "Ozone" column of this data frame?
```{r}
# mine
sum(sapply(dataset$Ozone, function(x) sum(is.na(x))))

# from the answer page:
miss <- is.na(dataset[, "Ozone"])  ## A vector of TRUE/FALSE
sum(miss)
```

9. What is the mean of the "Ozone" column in this dataset? Exclude missing values (coded as NA) from this calculation.
```{r}
# mine
mean(dataset$Ozone[!is.na(dataset$Ozone)])

# from the answer page:
mean(dataset[, "Ozone"], na.rm = TRUE)
```

10. Extract the subset of rows of the data frame where Ozone values are above 31 and Temp values are above 90.
```{r}
subset(dataset, Ozone > 31 & Temp > 90)
```

11. Use a for loop to create a vector of length 6 containing the mean of each column in the data frame (excluding all missing values).
```{r}
# mine
dataset.means <- vector()
for (col in colnames(dataset)) {
  dataset.means <- c(dataset.means,
                     mean(as.vector(dataset[[col]])[!is.na(dataset[[col]])]))
}

# from the answer page:
m <- numeric(6)
for (i in 1:6) {
    m[i] <- mean(dataset[, i], na.rm = TRUE)
}
print(m)
```

12. Use the apply function to calculate the standard deviation of each column in the data frame (excluding all missing values).
```{r}
# got a bit stuck here...
# minus the filtering bit...
# apply(dataset, colnames(dataset), sd)
# how to do this in a one-liner? w/o looping?

# from the answer page:
s <- apply(dataset, 2, sd, na.rm = TRUE)
print(s)

# Dave H. explains further:
# MARGIN = 2
for (i in 1:ncol(dataset)) {
  sd(dataset[, i], na.rm = TRUE)
}

# MARGIN = 1
for (i in 1:nrow(dataset)) {
  sd(dataset[i, ], na.rm = TRUE)
}
```

13. Calculate the mean of "Ozone" for each Month in the data frame and create a vector containing the monthly means (exclude all missing values).
```{r}
# ick ... loops are a code smell
dataset.ozone.means.by.month <- vector()
for (m in unique(dataset$Month)) {
  dataset.ozone.means.by.month <- c(dataset.ozone.means.by.month, 
                                    mean(subset(dataset, !is.na(Ozone) & Month==m)[["Ozone"]]))
}
# a way to do it without looping?

# from the answer page:
tapply(dataset$Ozone, dataset$Month, mean, na.rm = TRUE)

# Dave H. suggests:
library(plyr)
ddply(dataset, .(Month), function(d, ...) sd(d$Ozone, ...), na.rm = TRUE)
```

14. Draw a random sample of 5 rows from the data frame
```{r}
dataset[sample(nrow(dataset), 5), ]
```
