# [Data Analysis] Week 3 Lectures

## Example Analysis Assignment

[example file](https://d19vezwu8eufl6.cloudfront.net/dataanalysis/exampleProject.zip) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=83_en&format=txt) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=83)

"How did I do this analysis?"

* **prompt** - contains the question (naturally?)
* **assignment** - the materials that should be submitted
* **core** - has `finalcode` and `rawcode`
  * note that `raw`... contains _all_ analyses performed (not just the ones that appear in the final)
* **figures** - just the final figure
  * optionally (and/or: for yourself) consider keeping interesting exploratory figures
    * and/but - if you're using `Rmd` then you probably don't need to do this

## Exploratory Graphs (1 + 2)

slides
  [1](https://dl.dropbox.com/u/7710864/courseraPublic/week3/001exploratoryGraphs1/index.html) +
  [2](https://dl.dropbox.com/u/7710864/courseraPublic/week3/002exploratoryGraphs2/index.html) |
videos
  [1](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=85) +
  [2](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=87)

* "Why do we use graphs?"
  * _communication!_
  * see patterns that you can't see w/ numbers alone
  * understand properties of the data
  * suggest modeling strategies
* characteristics of _exploratory_ graphs
  * make many of them, and quickly
  * (clean it up later)
* (( illustrating how 1 set of data could be visualized _n_ different ways... ))
* "position vs. length"
  * ask: "What is the main take-away from the graph? Is it easy to 'see what you're saying'?"
  * (is the graph easy to read?)
* "position vs. angle"
  * ("why statisticians don't like pie charts")
* whenever possible: try to use **position** as the basis for graphical comparisons

### example plots
(using the American Community Survey data from last week's quiz)

```{r}
pData <- read.csv("../quizzes/ss06pid.csv")
```

#### Boxplot
* for a quantitative variable
* goal: get an idea of the distribution of the data
* observe:
  * **median** (thick line in the middle)
  * **75th** & **25th percentile** (upper & lower bounds of the box)
  * **whiskers** - 1.5x the value of 75th & 25th percentiles, respectively
  * also: outliers appear as dots
* `varwidth` = size the width based on the number of observations (for that factor)

```{r}
boxplot(pData$AGEP, col="blue")

# "Show me AGEP but broken down by DDRS"
boxplot(pData$AGEP ~ as.factor(pData$DDRS), col="blue")

# encode additional information with `varwidth`
boxplot(pData$AGEP ~ as.factor(pData$DDRS), col=c("blue", "orange"), names=c("yes", "no"), varwidth=TRUE)
```

#### Barplot
* height of the bar = data value
  * "number of observations per class"
* continuous data?
  * break it down into chunks and look at the values that way

```{r}
barplot(table(pData$CIT), col="blue")
```

#### Histograms
* important params: breaks, freq, col, xlab, ylab, xlim, ylim, main
* "sort of like boxplots"
  * goal : quanity a univariate distribution of data
* still chunking the distribution and summing them
  * (similar to the barplot example?)
  * and/but - more fine-grained
* helps to show the shape of the distribution
  * (in that regard: more resolution on the data than w/ a boxplot)
* you can set the number of breaks

```{r}
hist(pData$AGEP, col="blue", breaks=100)
```

#### Density Plots
* like a histogram but smoothed out
* warning: could introduce errors at the boundaries
  * careful when interpretting!

```{r}
dens <- density(pData$AGEP)
plot(dens, lwd=3, col="blue")
# `lwd` = 'line width' (roughly)

# useful (vs. histogram) b/c you can overlay them
dens <- density(pData$AGEP)
densMales <- density(pData$AGEP[which(pData$SEX=1)])
plot(dens, lwd=3, col="blue")
lines(densMales, lwd=3, col="orange")
```