[Data Analysis] Week 4 Lectures
================================================================================

**Preflight:**
```{r preflight}
setwd("~/Desktop/coursera-data-analysis/lectures/")
```

----

Clustering Example
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/courseraPublic/week4/001clusteringExample/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=97) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=97_en&format=txt)

Samsung Galaxy S3 case study / example

```{r}
load('samsungData.rda')
#names(samsungData)
table(samsungData$activity)

###
par(mfrow=c(1,2))
numericActivity <- as.numeric(as.factor(samsungData$activity))[samsungData$subject==1]
plot(samsungData[samsungData$subject==1,1],pch=19,col=numericActivity,ylab=names(samsungData)[1])
plot(samsungData[samsungData$subject==1,2],pch=19,col=numericActivity,ylab=names(samsungData)[2])
legend(150,-0.1,legend=unique(samsungData$activity),col=unique(numericActivity),pch=19)

### Clustering based _just on average_ acceleration
### cluster dendrogram (where are the patterns?)
source("http://dl.dropbox.com/u/7710864/courseraPublic/myplclust.R")
distanceMatrix <- dist(samsungData[samsungData$subject==1,1:3])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering,lab.col=numericActivity)

### Plotting **max** acceleration for the first subject
### (see the difference?)
par(mfrow=c(1,2))
plot(samsungData[samsungData$subject==1,10],
     pch=19, col=numericActivity,
     ylab=names(samsungData)[10])
plot(samsungData[samsungData$subject==1,11],
     pch=19, col=numericActivity,
     ylab=names(samsungData)[11])

### Clustering based on maximum acceleration
#source("http://dl.dropbox.com/u/7710864/courseraPublic/myplclust.R")
distanceMatrix <- dist(samsungData[samsungData$subject==1,10:12])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col=numericActivity)

### Singular value decomposition
svd1 = svd(scale(samsungData[samsungData$subject==1,-c(562,563)]))
par(mfrow=c(1,2))
plot(svd1$u[,1], col=numericActivity, pch=19)
plot(svd1$u[,2], col=numericActivity, pch=19)

### Find maximum contributor
### "look at the right singular vector that corresponds with the left singular vector that gives that pattern"
plot(svd1$v[,2], pch=19)

### New clustering with maximum contributer
maxContrib <- which.max(svd1$v[,2])
distanceMatrix <- dist(samsungData[samsungData$subject==1, c(10:12,maxContrib)])
hclustering <- hclust(distanceMatrix)
myplclust(hclustering, lab.col=numericActivity) 

### New clustering with maximum contributer
names(samsungData)[maxContrib]                          
### "separates out a variable that distinguishes between walking and walking-up"

### K-means clustering (nstart=1, first try)
kClust <- kmeans(samsungData[samsungData$subject==1,-c(562,563)],centers=6)
table(kClust$cluster,samsungData$activity[samsungData$subject==1])

### K-means clustering (nstart=1, second try)
### do it again! (note: not quite the same)
kClust <- kmeans(samsungData[samsungData$subject==1,-c(562,563)],centers=6,nstart=1)
table(kClust$cluster,samsungData$activity[samsungData$subject==1])

### K-means clustering (nstart=100, first try)
### 100 random starts, then take the average
kClust <- kmeans(samsungData[samsungData$subject==1,-c(562,563)],centers=6,nstart=100)
table(kClust$cluster,samsungData$activity[samsungData$subject==1])

### K-means clustering (nstart=100, second try)
kClust <- kmeans(samsungData[samsungData$subject==1,-c(562,563)],centers=6,nstart=100)
table(kClust$cluster,samsungData$activity[samsungData$subject==1])

### Cluster 1 Variable Centers (Laying)
plot(kClust$center[1,1:10],pch=19,ylab="Cluster Center",xlab="")

### Cluster 2 Variable Centers (Walking)
plot(kClust$center[6,1:10],pch=19,ylab="Cluster Center",xlab="")
```



----

Basic Least Squares
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/courseraPublic/week4/002basicLeastSquares/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=99) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=99_en&format=txt)

- 1st lecture about **regression**

### Goals of statistical modeling
- **Describe the distribution of variables**
- **Describe the relationship between variables**
- Make inferences about distributions or relationships

_Example: Average parent and child heights_

### The Galton Data

Load Galton Data:
```{r}
library(UsingR); data(galton)
par(mfrow=c(1,2))
hist(galton$child,col="blue",breaks=100)
hist(galton$parent,col="blue",breaks=100)
```

The distribution of child heights
```{r}
hist(galton$child,col="blue",breaks=100)
```

"If you had only one number to use to summarize the distribution what would it be?"

Only know the child - average height
```{r}
hist(galton$child,col="blue",breaks=100)
meanChild <- mean(galton$child)
lines(rep(meanChild,100),seq(0,150,length=100),col="red",lwd=5)
```

Here the mean makes sense b/c it's a pretty symmetric distribution

**Only know the child - why average?**

it minimizes a certain set of errors

What if we plot child versus average parent?
```{r}
plot(galton$parent,galton$child,pch=19,col="blue")
```

Jitter: (mostly just to visualize the points that are stacked)
```{r}
# Jittered plot
set.seed(1234)
plot(jitter(galton$parent,factor=2),jitter(galton$child,factor=2),pch=19,col="blue")
```

Average parent = 65 inches tall:
```{r}
plot(galton$parent,galton$child,pch=19,col="blue")
near65 <- galton[abs(galton$parent - 65)<1, ]
points(near65$parent,near65$child,pch=19,col="red")
lines(seq(64,66,length=100),rep(mean(near65$child),100),col="red",lwd=4)
```

"If the average parent is 65 inches tall, what is the average child height?"

Average parent = 71 inches tall:
```{r}
plot(galton$parent,galton$child,pch=19,col="blue")
near71 <- galton[abs(galton$parent - 71)<1, ]
points(near71$parent,near71$child,pch=19,col="red")
lines(seq(70,72,length=100),rep(mean(near71$child),100),col="red",lwd=4)
```

Fitting a line
```{r}
plot(galton$parent,galton$child,pch=19,col="blue")
lm1 <- lm(galton$child ~ galton$parent)
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```

**linear model**

_"Why not this line?"_
```{r}
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent, 26 + 0.646*galton$parent)
```

B/c *Not all points are on the line*
```{r}
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```

Allowing for variation (e.g., what are the variables we **didn't** measure?)

**How do we pick best?**

> If $C_i$ is the height of child $i$ and $P_i$ is the height of the average parent,
> pick the line that makes the child values $C_i$ and our guesses

**Plot what is leftover**
```{r}
par(mfrow=c(1,2))
plot(galton$parent,galton$child,pch=19,col="blue")
lines(galton$parent,lm1$fitted,col="red",lwd=3)
plot(galton$parent,lm1$residuals,col="blue",pch=19)
abline(c(0,0),col="red",lwd=3)
```



----

Inference Basics
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/courseraPublic/week4/003inferenceBasics/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=101) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=101_en&format=txt)

- prev'ly: fitting lines
- if we have a _sample_ - how can we say something about our best fit line?
  - what does that line say about the _population_?
  
### Fit a line to the Galton Data
```{r}
library(UsingR)
data(galton)
plot(galton$parent,galton$child,pch=19,col="blue")
lm1 <- lm(galton$child ~ galton$parent)
lines(galton$parent,lm1$fitted,col="red",lwd=3)
```

- plot average parent height vs. child height
- `lm` = "linear model" fn
  - `lm` always fits an intercept term (unless you tell it not to)
- use `lines` to add the fit line (From `lm`) to the `plot`

```{r}
lm1
```

- observe the coefficient(s)!

### Create a "population" of 1 million families
- "Create a scenario where we have 1 million families..."
  - based on the fit of the original data set

```{r}
newGalton <- data.frame(parent=rep(NA, 1e6), child=rep(NA,1e6))

# create the parents (use mean & sd of parents from Galton data)
# again: `rnorm` to generate this normally distributed random sample
newGalton$parent <- rnorm(1e6,mean=mean(galton$parent), sd=sd(galton$parent))

# create the children from the parent data and the coefficients from the `lm`
newGalton$child <- lm1$coeff[1] + lm1$coeff[2]*newGalton$parent + rnorm(1e6,sd=sd(lm1$residuals))

smoothScatter(newGalton$parent,newGalton$child)
abline(lm1,col="red",lwd=3)
# note how the new, simulated data actually fits to our regression line from the actual Galton data
```

### Let's take a sample
```{r}
set.seed(134325)

# sample from our simulated data
sampleGalton1 <- newGalton[sample(1:1e6,size=50,replace=FALSE),]

# new regression line (from the simulated/random sample)
sampleLm1 <- lm(sampleGalton1$child ~ sampleGalton1$parent)

plot(sampleGalton1$parent,sampleGalton1$child,pch=19,col="blue")

# put our black (sample) & red (actual) lines on there
lines(sampleGalton1$parent,sampleLm1$fitted,lwd=3,lty=2)
abline(lm1,col="red",lwd=3)
```

### Let's take another sample
"Another random sample of 50..."

```{r}
sampleGalton2 <- newGalton[sample(1:1e6,size=50,replace=FALSE),]
sampleLm2 <- lm(sampleGalton2$child ~ sampleGalton2$parent)
plot(sampleGalton2$parent,sampleGalton2$child,pch=19,col="blue")
lines(sampleGalton2$parent,sampleLm2$fitted,lwd=3,lty=2)
abline(lm1,col="red",lwd=3)
```

- again: red line is the original fit line (real Galton data)
- black line is our random sample
  - sample is always random, so fit line is always a little different
  
### Let's take another sample
Again:
```{r}
sampleGalton3 <- newGalton[sample(1:1e6,size=50,replace=F),]
sampleLm3 <- lm(sampleGalton3$child ~ sampleGalton3$parent)
plot(sampleGalton3$parent,sampleGalton3$child,pch=19,col="blue")
lines(sampleGalton3$parent,sampleLm3$fitted,lwd=3,lty=2)
abline(lm1,col="red",lwd=3)
```

### Many samples
"Let's do that again. 100 times."
```{r}
sampleLm <- vector(100,mode="list")
for(i in 1:100){
  sampleGalton <- newGalton[sample(1:1e6,size=50,replace=FALSE),]
  sampleLm[[i]] <- lm(sampleGalton$child ~ sampleGalton$parent)
}

smoothScatter(newGalton$parent,newGalton$child)
for(i in 1:100){abline(sampleLm[[i]],lwd=3,lty=2)}
abline(lm1,col="red",lwd=3)
```
- again: red line is original fit line from real data
- the many black lines? - fit lines from the 100 iterations
  - note the different slopes etc.
  - and/but mostly around the same center
- real thing? - you only get one of these (black) lines in a given study

### Histogram of estimates
```{r}
par(mfrow=c(1,2))
# hist. of intercept terms:
hist(sapply(sampleLm,function(x){coef(x)[1]}),col="blue",xlab="Intercept",main="")
# hist. of slope terms:
hist(sapply(sampleLm,function(x){coef(x)[2]}),col="blue",xlab="Slope",main="")
```
- In real world you don't get to look at this many.
- in the real world you usually only have one sample
  - and you need to work w/ what you have

### Distribution of coefficients
- **central limit theorem** - _super important_
- re: estimating variance
- **standard error** = square root of the estimated variance for a coefficient

### Estimating the values in R
```{r}
par(mfrow=c(1,1))
sampleGalton4 <- newGalton[sample(1:1e6,size=50,replace=FALSE),]
sampleLm4 <- lm(sampleGalton4$child ~ sampleGalton4$parent)
summary(sampleLm4)

hist(sapply(sampleLm,function(x){coef(x)[2]}),col="blue",xlab="Slope",main="",freq=F)
lines(seq(0,5,length=100),dnorm(seq(0,5,length=100),mean=coef(sampleLm4)[2],
      sd=summary(sampleLm4)$coeff[2,2]),lwd=3,col="red")
```
- note actual slope is slightly different than the slope value that generated the data
  - but the variance is about the same

### Why do we standardize?
- **standardize!**
  - b/c you want to be able to make comparisons
  - and standardized quantitative data is easier to compare
  - easier to reason about
```{r}
par(mfrow=c(1,2))
hist(sapply(sampleLm,function(x){coef(x)[1]}),col="blue",xlab="Intercept",main="")
hist(sapply(sampleLm,function(x){coef(x)[2]}),col="blue",xlab="Slope",main="")
```

### Standardized coefficients
- take the coefficients that we have...
- Degrees of Freedom: (approx.)
  - number of samples - number of things you estimated
  
### $t_{n−2}$ versus $N(0,1)$
```{r}
par(mfrow=c(1,1))

# linear set of points
x <- seq(-5,5,length=100)
# plot the normal dist.:
plot(x,dnorm(x),type="l",lwd=3)
# and a couple of t-distributions:
lines(x,dt(x,df=3),lwd=3,col="red")
lines(x,dt(x,df=10),lwd=3,col="blue")
```

### Confidence intervals
- "what are the possible values that the real B1 has?"
- level $\alpha$ confidence interval
  - confidence interval: set of values for B1 that we think are plausible
```{r}
summary(sampleLm4)$coeff
confint(sampleLm4,level=0.95)
```
- note on the slide: "+" and "-" are backward in the equation

Repeated samples of size 50 from the 1 million family data set:
```{r}
par(mar=c(4,4,0,2))
plot(1:10,type="n",xlim=c(0,1.5),ylim=c(0,100),
     xlab="Coefficient Values",ylab="Replication")
for(i in 1:100){
    ci <- confint(sampleLm[[i]])
    color="red"
    if((ci[2,1] < lm1$coeff[2]) & (lm1$coeff[2] < ci[2,2])){color = "grey"}
    segments(ci[2,1],i,ci[2,2],i,col=color,lwd=3)
}
lines(rep(lm1$coeff[2],100),seq(0,100,length=100),lwd=3)
```

### How you report the inference
```{r}
sampleLm4$coeff
confint(sampleLm4,level=0.95)
```

> A one inch increase in parental height is associated with a 0.77 inch
> increase in child's height (95% CI: 0.42-1.12 inches).



----

P-values
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/courseraPublic/week4/004pValues/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=105) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=105_en&format=txt)

- Most common measure of "statistical significance"
- Commonly reported in papers
- Used for decision making (e.g. FDA)
- Controversial among statisticians
  - <http://warnercnr.colostate.edu/~anderson/thompson1.html>
  - primary reason: widely used but also widely misinterpretted
- **Not everyone thinks P-values are awful**
  - <http://simplystatistics.org/2012/01/06/p-values-and-hypothesis-testing-get-a-bad-rap-but-we/>

### What is a P-value?
**Idea:** Suppose nothing is going on - _how unusual is it to see the estimate we got?_

**Approach:**
1. Define the hypothetical distribution of a data summary (statistic) when "nothing is going on" (_null hypothesis_)
2. Calculate the summary/statistic with the data we have (_test statistic_)
3. Compare what we calculated to our hypothetical distribution and see if the value is "extreme" (_p-value_)

### Galton data
(again; case study cont'd)

```{r}
library(UsingR)
data(galton)
plot(galton$parent, galton$child,
     pch=19, col="blue")
lm1 <- lm(galton$child ~ galton$parent)
abline(lm1,col="red",lwd=3)
```
- **null hypothesis** - no relationship b/w parent + child height
- **null dist.** + **observed stat.**
```{r}
x <- seq(-20,20,length=100)
plot(x,dt(x,df=(928-2)),col="blue",lwd=3,type="l")
arrows(summary(lm1)$coeff[2,3],0.25,summary(lm1)$coeff[2,3],0,col="red",lwd=4)
```

### Calculating p-values
```{r}
summary(lm1)
```

### A quick simulated example
```{r}
set.seed(9898324)
yValues <- rnorm(10)
xValues <- rnorm(10)
lm2 <- lm(yValues ~ xValues)
summary(lm2)
```
- random `x` and `y`
  - should be totally unrelated
```{r}
x <- seq(-5,5,length=100)
plot(x,dt(x,df=(10-2)),col="blue",lwd=3,type="l")
arrows(summary(lm2)$coeff[2,3],0.25,summary(lm2)$coeff[2,3],0,col="red",lwd=4)
```
- "How much probability density exists to the _right_ of the observed statistic?"
  - how much curve lies to the right of the red arrow?
```{r}
xCoords <- seq(-5,5,length=100)
plot(xCoords,dt(xCoords,df=(10-2)),col="blue",lwd=3,type="l")
xSequence <- c(seq(summary(lm2)$coeff[2,3],5,length=10),summary(lm2)$coeff[2,3])
ySequence <- c(dt(seq(summary(lm2)$coeff[2,3],5,length=10),df=8),0)
polygon(xSequence,ySequence,col="red"); polygon(-xSequence,ySequence,col="red")
```

### Simulate a ton of data sets with no signal
```{r}
set.seed(8323); pValues <- rep(NA,100)
for(i in 1:100){
  xValues <- rnorm(20);yValues <- rnorm(20)
  pValues[i] <- summary(lm(yValues ~ xValues))$coeff[2,4]
}
hist(pValues,col="blue",main="",freq=F)
abline(h=1,col="red",lwd=3)

set.seed(8323); pValues <- rep(NA,100)
for(i in 1:100){
  xValues <- rnorm(20);yValues <- 0.2 * xValues + rnorm(20)
  pValues[i] <- summary(lm(yValues ~ xValues))$coeff[2,4]
}
hist(pValues,col="blue",main="",freq=F,xlim=c(0,1)); abline(h=1,col="red",lwd=3)
```

> "Under the null hypothesis, the p-values will always be uniformly distributed."
> "If the alternative hypothesis is true..."

```{r}
set.seed(8323); pValues <- rep(NA,100)
for(i in 1:100){
  xValues <- rnorm(100);yValues <- 0.2* xValues + rnorm(100)
  pValues[i] <- summary(lm(yValues ~ xValues))$coeff[2,4]
}
hist(pValues,col="blue",main="",freq=F,xlim=c(0,1)); abline(h=1,col="red",lwd=3)
```
- p-value depends on the signal (the slope) and the sample size

### Some typical values (single test)
- $P < 0.05$ (significant)
- $P < 0.01$ (strongly significant)
- $P < 0.001$ (very significant)

> In modern analyses, people generally report both the confidence interval and
> P-value. This is less true if many many hypotheses are tested.

If you are working w/ lots of data, try to use a P-value smaller than 0.05.

### How you interpret the results
```{r}
summary(lm(galton$child ~ galton$parent))$coeff
```

> A one inch increase in parental height is associated with a 0.77 inch
> increase in child's height (95% CI: 0.42-1.12 inches). This difference was
> statistically significant (P<0.001).

P-value on its own is not necessarily super-informative but can be useful
in guiding "where to from here" etc.



----

Regression with Factor Variables
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/courseraPublic/week4/005factorVariables/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=107) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=107_en&format=txt)

Key ideas:
- Outcome is still quantitative
- Covariate(s) are factor variables
  - "factor" roughly equal "qualitative"
- Fitting lines = fitting means
- Want to evaluate contribution of all factor levels at once

Example: movie ratings

Go get some RottenTomatoes.com movie ratings:

```{r}
download.file("http://www.rossmanchance.com/iscam2/data/movies03RT.txt",destfile="data/movies.txt")
movies <- read.table("data/movies.txt",sep="\t",header=T,quote="")
head(movies)
```

### Rotten tomatoes score vs. rating
```{r}
plot(movies$score ~ jitter(as.numeric(movies$rating)),col="blue",xaxt="n",pch=19)
axis(side=1,at=unique(as.numeric(movies$rating)),labels=unique(movies$rating))
```
Get the ratings and plot them by their rating.

### Average score by rating
```{r}
plot(movies$score ~ jitter(as.numeric(movies$rating)),col="blue",xaxt="n",pch=19)
axis(side=1,at=unique(as.numeric(movies$rating)),labels=unique(movies$rating))
meanRatings <- tapply(movies$score,movies$rating,mean)
points(1:4,meanRatings,col="red",pch="-",cex=5)
```
Apply a marker to indicate the mean score per rating.

### Another way to write it down
(( insert mathematical equation here... ))

**Average values**
- $b_0$ = average of the G movies
- $b_0 + b_1$ = average of the PG movies
- $b_0 + b_2$ = average of the PG-13 movies
- $b_0 + b_3$ = average of the R movies

- slightly more complicated than just calc'ing the averages
- BUT: this allows us to fit them w/ linear models

### Doing it in R
```{r}
lm1 <- lm(movies$score ~ as.factor(movies$rating))
summary(lm1)
```
Note: coercing the ratings to factors w/ `as.factor`

### Plot fitted values
```{r}
plot(movies$score ~ jitter(as.numeric(movies$rating)),col="blue",xaxt="n",pch=19)
axis(side=1,at=unique(as.numeric(movies$rating)),labels=unique(movies$rating))
points(1:4,lm1$coeff[1] + c(0,lm1$coeff[2:4]),col="red",pch="-",cex=5)
```

### Question 1
"What is the average difference in rating b/w G- & R-rated movies?"

```{r}
lm1 <- lm(movies$score ~ as.factor(movies$rating))
summary(lm1)
confint(lm1)
```

### Question 2
"What is the average difference in rating b/w PG-13 and R-rated movies?"

_We could rewrite our model..._

```{r}
# `ref` is the "reference category"
lm2 <- lm(movies$score ~ relevel(movies$rating,ref="R"))
summary(lm2)
confint(lm2)
```

### Question 3
"What if you want to know if there is any difference at all b/w any of the
different levels of this particular variable?" (at any level?)

ANOVA!

```{r}
lm1 <- lm(movies$score ~ as.factor(movies$rating))
anova(lm1)
```

### Sum of squares (G movies)
(remember? from ANOVA (above))

```{r}
gMovies <- movies[movies$rating=="G",]; xVals <- seq(0.2,0.8,length=4)
plot(xVals,gMovies$score,ylab="Score",xaxt="n",xlim=c(0,1),pch=19)
abline(h=mean(gMovies$score),col="blue",lwd=3); abline(h=mean(movies$score),col="red",lwd=3)
segments(xVals+0.01,rep(mean(gMovies$score),length(xVals)),xVals+0.01,
         rep(mean(movies$score),length(xVals)),col="red",lwd=2)
segments(xVals-0.01,gMovies$score,xVals-0.01,rep(mean(gMovies$score),length(xVals)),col="blue",lwd=2)
```

### Tukey's (honestly significant difference test)
less common than the ANOVA
```{r}
lm1 <- aov(movies$score ~ as.factor(movies$rating))
TukeyHSD(lm1)
```



----

Multiple Variable Regression
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/courseraPublic/week4/006multipleVariables/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=103) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=103_en&format=txt)

"Same least squares approach..."

**Key ideas:**
- Regression with multiple covariates
- Still using least squares/central limit theorem
- Interpretation depends on all variables

Example data: Millenium Development Goal 1 + WHO childhood hunger

```{r}
download.file("http://apps.who.int/gho/athena/data/GHO/WHOSIS_000008.csv?profile=text&filter=COUNTRY:*;SEX:*","data/hunger.csv",method="curl")
hunger <- read.csv("data/hunger.csv")
hunger <- hunger[hunger$Sex!="Both sexes",]
head(hunger)
```

### Plot percent hungry versus time
```{r}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
```

**"Remember the linear model"**

$$ Hu_i = b_0 + b_1Y_i + e_i $$

- $b_0$ = percent hungry at Year 0
- $b_1$ = decrease in percent hungry per year
- $e_i$ = everything we didn't measure

### Add the linear model
```{r}
lm1 <- lm(hunger$Numeric ~ hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19,col="blue")
lines(hunger$Year,lm1$fitted,lwd=3,col="darkgrey")
```

### Color by male/female
```{r}
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
```

**"Now two lines..."*** (Male/Female w/ separate intercept lines)
```{r}
lmM <- lm(hunger$Numeric[hunger$Sex=="Male"] ~ hunger$Year[hunger$Sex=="Male"])
lmF <- lm(hunger$Numeric[hunger$Sex=="Female"] ~ hunger$Year[hunger$Sex=="Female"])
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
lines(hunger$Year[hunger$Sex=="Male"],lmM$fitted,col="black",lwd=3)
lines(hunger$Year[hunger$Sex=="Female"],lmF$fitted,col="red",lwd=3)
```

### Two lines, same slope
(in R)
```{r}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] ),col="black",lwd=3)
```

### Interactions
**Two lines, different slopes**
(in R)
```{r}
lmBoth <- lm(hunger$Numeric ~ hunger$Year + hunger$Sex + hunger$Sex*hunger$Year)
plot(hunger$Year,hunger$Numeric,pch=19)
points(hunger$Year,hunger$Numeric,pch=19,col=((hunger$Sex=="Male")*1+1))
abline(c(lmBoth$coeff[1],lmBoth$coeff[2]),col="red",lwd=3)
abline(c(lmBoth$coeff[1] + lmBoth$coeff[3],lmBoth$coeff[2] +lmBoth$coeff[4]),col="black",lwd=3)

summary(lmBoth)
```

### Interactions for continuous variables
"a little bit tricky" (!?)



----

Regression in the Real World
--------------------------------------------------------------------------------

[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=109) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=109_en&format=txt)

(( just watch ))

- "Previous examples ... everything works out neatly and nicely."
  - "In real life ... regression is very hard" (complicated)
- **Things to pay attn to:**
  - _confounders_
    - variable that is correlated with both dependent & independent variables
    - sometimes detect w/ careful exploration/visualization
  - complicated interactions
  - skewness
  - _outliers_
    - data points that are outside the expectations
      - ("way too high or way too low")
      - ("something that doesn't fit the pattern")
    - even 1 outlier can dramatically alter the "shape of the data
      - and/or the visualization
    - careful w/ outliers
      - sometimes they're b/c of broken data collection
      - and/but sometimes they're real data
        - (that just happen to introduce big-time skew...)
  - non-linear patterns
  - _variance changes_
    - heteroskedasticity
    - there are some things you can do to deal w/ this
      - (list of models/techniques)
  - _issues w/ diff. units or scales_
    - sometimes you need to standardize
      - not just ensuring they're all using the same units
    - also: "flatten" some of the data (e.g., not just "total deaths" but "deaths per 1000" or some such thing)
  - overloading regression (??)
  - correlation & causation
- Ideal data for regression?
  - oval shape (like Galton's)




----

Study Group Supplementary Session
--------------------------------------------------------------------------------

**Dave explains residuals from a linear model:**

$$ y = mx + b + e $$

- Dave: "simplest linear model" (above)
  - $mx$ = slow
  - $b$ = intercept
- residual = $r_i = y_i - y(hat)_i$
  - $= y_i - (mx_i + b)$
  - difference b/w the prediction made by the model and the _actual observed value_
- _coefficients_ are the values that go into the model to make the predictions
  - so: $m$ and $b$ in our "simplest model" (above)