# [Data Analysis] Week 3 Lectures

## Example Analysis Assignment

[example file](https://d19vezwu8eufl6.cloudfront.net/dataanalysis/exampleProject.zip) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=83_en&format=txt) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=83)

"How did I do this analysis?"

* **prompt** - contains the question (naturally?)
* **assignment** - the materials that should be submitted
* **core** - has `finalcode` and `rawcode`
  * note that `raw`... contains _all_ analyses performed (not just the ones that appear in the final)
* **figures** - just the final figure
  * optionally (and/or: for yourself) consider keeping interesting exploratory figures
    * and/but - if you're using `Rmd` then you probably don't need to do this

## Exploratory Graphs (1 + 2)

slides
  [1](https://dl.dropbox.com/u/7710864/courseraPublic/week3/001exploratoryGraphs1/index.html) +
  [2](https://dl.dropbox.com/u/7710864/courseraPublic/week3/002exploratoryGraphs2/index.html) |
videos
  [1](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=85) +
  [2](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=87)

* "Why do we use graphs?"
  * _communication!_
  * see patterns that you can't see w/ numbers alone
  * understand properties of the data
  * suggest modeling strategies
* characteristics of _exploratory_ graphs
  * make many of them, and quickly
  * (clean it up later)
* (( illustrating how 1 set of data could be visualized _n_ different ways... ))
* "position vs. length"
  * ask: "What is the main take-away from the graph? Is it easy to 'see what you're saying'?"
  * (is the graph easy to read?)
* "position vs. angle"
  * ("why statisticians don't like pie charts")
* whenever possible: try to use **position** as the basis for graphical comparisons

### example plots
(using the American Community Survey data from last week's quiz)

```{r}
pData <- read.csv("../quizzes/ss06pid.csv")
```

#### Boxplot
* for a quantitative variable
* goal: get an idea of the distribution of the data
* observe:
  * **median** (thick line in the middle)
  * **75th** & **25th percentile** (upper & lower bounds of the box)
  * **whiskers** - 1.5x the value of 75th & 25th percentiles, respectively
  * also: outliers appear as dots
* `varwidth` = size the width based on the number of observations (for that factor)

```{r}
boxplot(pData$AGEP, col="blue")

# "Show me AGEP but broken down by DDRS"
boxplot(pData$AGEP ~ as.factor(pData$DDRS), col="blue")

# encode additional information with `varwidth`
boxplot(pData$AGEP ~ as.factor(pData$DDRS), col=c("blue", "orange"), names=c("yes", "no"), varwidth=TRUE)
```

#### Barplot
* height of the bar = data value
  * "number of observations per class"
* continuous data?
  * break it down into chunks and look at the values that way

```{r}
barplot(table(pData$CIT), col="blue")
```

#### Histograms
* important params: breaks, freq, col, xlab, ylab, xlim, ylim, main
* "sort of like boxplots"
  * goal : quanity a univariate distribution of data
* still chunking the distribution and summing them
  * (similar to the barplot example?)
  * and/but - more fine-grained
* helps to show the shape of the distribution
  * (in that regard: more resolution on the data than w/ a boxplot)
* you can set the number of breaks

```{r}
hist(pData$AGEP, col="blue", breaks=100)
```

#### Density Plots
* like a histogram but smoothed out
* warning: could introduce errors at the boundaries
  * careful when interpretting!

```{r}
dens <- density(pData$AGEP)
plot(dens, lwd=3, col="blue")
# `lwd` = 'line width' (roughly)

# useful (vs. histogram) b/c you can overlay them
dens <- density(pData$AGEP)
densMales <- density(pData$AGEP[which(pData$SEX==1)])
plot(dens, lwd=3, col="blue")
lines(densMales, lwd=3, col="orange")
```

#### Scatterplots
* most widely used plot for exploratory analysis
* `x` & `y` are critical variables
  * (list of several other important params)
  * `?par` for more
* remember: each point represents 1 observation
* interesting: scatterplots can help illuminate weird patterns quickly
  * and/or outliers become flaringly obvious
* "size matters"
  * pay attn to scale on axis & size of dots

```{r}
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue")

# try again:
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue", cex=0.5)

# w/ color: (remember: need a number value for `col`)
plot(pData$JWMNP, pData$WAGP, pch=19, col=pData$SEX, cex=0.5)

# you can also use size to illustrate "the third variable"
percentMaxAge <- pData$AGEP/max(pData$AGEP)
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue", cex=percentMaxAge*0.5)

# get fancy: overlay lines/points
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue", cex=0.5)
lines(rep(100, dim(pData)[1]), pData$WAGP, col="grey", lwd=5)
points(seq(0, 200, length=100), seq(0, 20e5, length=100), col="red", pch=19)

# more fancy: visualizing numeric variables as factors
library(Hmisc)
ageGroups <- cut2(pData$AGEP, g=5) # <-- breaks up the age groups into "factors" (5 chunks of age groups)
plot(pData$JWMNP, pData$AGEP, pch=19, col=ageGroups, cex=0.5)
```

"If you have lots of points..." (e.g., 100k)

```{r}
# big black mass!
x <- rnorm(1e5)
y <- rnorm(1e5)
plot(x, y, pch=19)

# try sampling the values
x <- rnorm(1e5)
y <- rnorm(1e5)
sampledValues <- sample(1:1e5, size=1000, replace=FALSE)
plot(x[sampledValues], y[sampledValues], pch=19)

# ...or a "smooth scatter"
x <- rnorm(1e5)
y <- rnorm(1e5)
smoothScatter(x,y)

# ...or "hexbin"
library(hexbin)
x <- rnorm(1e5)
y <- rnorm(1e5)
hbo <- hexbin(x,y)
plot(hbo)
```

#### QQ Plots

* "Sort of like a scatter plot but with a very particular purpose..."
* plots quantiles of `x` vs. quantiles of `y`

```{r}
x <- rnorm(20)
y <- rnorm(20)
qqplot(x,y)
abline(c(0,1))
```

#### Matplot and spaghetti

* takes each column and plots it as one line
  * each column (in the matrix) becomes one specific line
* compare trends/trajectories over time

```{r}
X <- matrix(rnorm(20*5), nrow=20)
matplot(X, type="b")
```

#### Heatmaps

* "sort of like a 2D histogram"
* goal: visualize the whole matrix of data
* color --> intensity

```{r}
image(1:10, 161:236, as.matrix(pData[1:10, 161:236]))

# might want to transpose that...
newMatrix <- as.matrix(pData[1:10, 161:236])
newMatrix <- t(newMatrix)[,nrow(newMatrix):1]
image(161:236, 1:10, newMatrix)
```

#### Maps (very basics)

* very basics
  * too much to cover
* use lat/lon values w/ these

```{r}
# let's get a map
library(maps)
map("world")

# let's make up some stuff and throw it on the map
lat <- runif(40, -180, 180)
lon <- runif(40, -90, 90)
points(lat, lon, col="blue", pch=19)
```

#### Missing values and plots

* can use plots to help understand what values are missing
* gaps in the plot? check the data for `NA`!
* and/but/also : try using the `boxplot`
  * relationships b/w missing values and... values?

### Further Resources

* [R Graph Gallery](http://gallery.r-enthusiasts.com/)
* [ggplot2](http://cran.r-project.org/web/packages/ggplot2/index.html)
  * [basic introduction](http://www.r-bloggers.com/basic-introduction-to-ggplot2/)
* [lattice](http://cran.r-project.org/web/packages/lattice/index.html)
  * [introduction](http://lmdvr.r-forge.r-project.org/figures/figures.html)
* [R bloggers](http://www.r-bloggers.com/)
