[Data Analysis] Week 1 Lecture Notes
================================================================================

Course Introduction
--------------------------------------------------------------------------------

[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=7) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=7_en&format=txt)

- Jeff Leeks - professor, biostatistics at JHU 
  - [Simply Statistics blog](http://simplystatistics.org/)
- (( going to be pretty comprehensive, but not enough time to go in-depth on it all ))



----

Getting Help
--------------------------------------------------------------------------------

[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=9) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=9_en&format=txt) |
[slides](https://dl.dropbox.com/u/7710864/002gettingHelp/index.html)

- *where to get help?*
  - message boards are one place... to ask questions 
- *where to look for answers?*
  - ( before asking the question to the boards... )
  - R questions?
    - [Roger Peng's video lecture series](http://www.youtube.com/watch?v=ZFaWxxzouCY&list=PLjTlxb-wKvXNSDfcKPFH2gzHGyjpeCZmJ&index=3)
    - search Roger's R class forums 
    - [R mailing list](http://www.r-project.org/mail.html)
    - search for R on Stack Overflow 
   - stats questions?
     - forums...
     - [CrossValidated](http://stats.stackexchange.com/) 
- *"Googling..."*
  - CrossValidated for general data analysis questions 
  - "half the hurdle" → what do you call "data analysis" in your given field?
    - e.g., data analysis? or econometrics? or biostatistics? 



----

What _is_ Data?
--------------------------------------------------------------------------------

[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=45) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=45_en&format=txt) |
[slides](https://dl.dropbox.com/u/7710864/003whatIsData/index.html)

- (( start w/ the Wikipedia definition )) → and/but break it down 
  - Data are values of qualitative or quantitative variables, belonging to a set of items.
- **set of items** → population or sample 
- **variables** → characteristic of what you're measuring 
  - qualitative (categorical) or quantitative (numerical) 
- **raw data** → "original source" 
  - usually hard to manipulate 
- **processed data** → data that is ready for analysis 
  - (after you've sliced/diced it)
  - e.g., normalizing values; dealing with outliers; transforming sentinel values; eliminating values that are of no interest...
- ( example of pipeline b/w raw and processed data )
  - the gene sequencing case study (what is the "raw data"? the image? the intensity values? the "raw" summary?)
- processed data → what does it look like?
  - clean columns, clean rows (tables, data frames) 
- "What is big data?"
  - "Any data that is too big to analyze with a single computer."
  - ( but that's easy to dispute depending on any number of factors )
  - "so easy to collect so much data in so little time and for so little money"



----

Representing Data
--------------------------------------------------------------------------------

[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=49) |
[slides](https://dl.dropbox.com/u/7710864/004representingData/index.html)

- "How do we represent data?" 
  - important to know how to write about it → b/c _The Point_ is communicating 
- How do we write about data?
  - each data point gets a capital letter
    - e.g., $H$ for height 
    - sub-script numerals if you have more than one for that type
  - $Y$ generally represents outcomes, $X$ for covariates
- **Randomness**
  - $X$ & $Y$ as "random variables" 
  - a variable is random b/c it is an incompletely measured variable 
    - or it's a random draw from a population 
  - $X = x$ → fixing $X$ (random quantity $X$) to be observed/fixed $x$
- example : coin flip (canonical "random") 
  - and/but so : studies indicate that even the coin flip can be deterministic 
    - so when you're marking something as random it's b/c you don't have the data to make a claim about it in a deterministic fashion 
- **Distributions**
  - $Pr$ = probability (b/w 0 and 1, must add up to 1)
  - our coin flip probability written as:
    - $Pr(X = 1) = 0.5$ and $Pr(X = 0) = 0.5$
      - in other words: 50/50 chance of heads or tails
- continuous vs. discrete distributions 
  - discrete → qualitative 
  - continuous → quantitative (ranges) 
- parameters 
  - "Distributions are defined by a set of fixed values called _parameters_."
  - usually represented w/ Greek letters 
    - $N(\mu,\sigma)$ or $Poisson(\lambda)$
  - e.g., $X ~ N(\mu,\sigma)$
    - translation: $X$ has the $N(\mu,\sigma)$ distribution 
- 3 most important parameters 
  - $E[X]$ → expectation of X (measures the "center"; usually mean?) 
  - $Var[X]$ → variance of X 
  - $SD[X]$ → standard deviation 
- **Conditioning**
  - $X|\mu$ → X is a random variable with fixed parameter $\mu$
- example distribution : **Binomial** 
  - coin flips:
  - $Bin(n, p)$
  - $X ~ Bin(10, 0.5)$
- example distribution : **Normal**
  - $N(\mu,\sigma)$
    - $\mu$ = mean
    - $\sigma$ = standard deviation
  - $X ~ N(0, 1)$
  - ( "probability density function" )
- example distribution : **Uniform**
  - "all values are equally likely" 
  - $U(\alpha, \beta)$
    - $\alpha, \beta$ = boundaries of which values we consider to be equally likely
  - $X ~ U(0, 1)$
- **Changing parameters**
  - Normal distribution:
    - $X ~ N(0, 1) , E[X] = \mu = 0, Var[X] = \sigma^2 = 1$
      - vs. 
    - $X ~ N(0, 5) , E[X] = \mu = 0, Var[X] = \sigma^2 = 25$
      - (a fatter "bell curve" function)
      - vs. 
    - $X ~ N(5, 1) , E[X] = \mu = 5, Var[X] = \sigma^2 = 1$
      - (skewed)
  - Binomial:
    - $X ~ Bin(10, 0.5), E[X] = n \times p = 5, Var[X] = n \times p (1 - p) = 2.5$
      - vs.
    - $X ~ Bin(10, 0.8), E[X] = n \times p = 8, Var[X] = n \times p (1 - p) = 1.6$
      - (skewed)
  - CONDITION:
    - Suppose : $Y ~ N(X, 1)$ and $X ~ N(0, 1)$, then distribution of $Y|X = 5$
    - (( conditioning can help reduce some of the variance b/c it holds one of the random variables constant, meaning that the function does NOT need to account for the variance of both (just the one) ))



----

Representing Data in R
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/005representingDataR/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=51) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=51_en&format=txt)

- (( more/less an intro to R and its core concepts ))
  - recommended video: [Peng's "Lecture 2a: Data Types" (video)](http://www.youtube.com/watch?v=5AQM-yUX9zg)
- data types:
  - `character` ("Jeff") 
  - `numeric` (1) 
  - `integer` (1L) 
  - `logical` (`TRUE` or `FALSE`) 
  - `vectors` (get familiar w/ these...; homogenous types)
  - `lists` (like vectors, but heterogeneous)
  - `matrices` (you know... like from math class)
  - data frames (get familiar w/ these → most of our data will be in data frames)
    - think : "like a spreadsheet" (columns & row)
  - **factors** → "Qualitative variables that can be included in models"
    - `as.factor`
    - very important!! 
  - missing values (`NA`)
- subsetting 
  - bracket notation
    - `vector[i]` → where i is an index
    - `vector[c(1:5)]` → where you give the brackets a vector of the indices (plural) you want to get back out 
  - bracket notation for specific rows, columns in a data frame 
    - `dataframe[1, 2:5]`
      - `[row(s), column(s)]`
  - `$dollarSignOperator` 
    - `my.data$columnName`
  - more bracket notation → with an expression for filtering
    - `my.data[heights < 190, ]`
- variable naming conventions 
  - acceptable: camelCase, with_underscores
  - idiomatic: with.dots 
- Style guides are out there for R:
  - e.g., [Google's R style guide](http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html)



----

Simulation Basics
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/006simulationBasics/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=53) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=53_en&format=txt)

- basics of simulation →
  - primarily (at beginning of class) for generating idealized data sets 
- 3 categories:
  - distributions
    - all functions start with `r`
  - densities
    - all functions start with `d`
  - sampling
- "r functions" 
  - to generate data → example:
```{r}
coin.flips <- rbinom(10, size=10, prob=0.5)
```
- "d functions"
  - to generate data → example:
  - (see also : `dbinom`) 
```{r}
x <- seq(from=-5, to=5, length=10)
normal.density <- dnorm(x, mean=0, sd=1)
round(normal.density, 2)
```
- sample → to sample from data we already have
  - see also : prob arg for probabilities (takes a vector of probabilities (remember: probabilities must sum to 1))
```{r}
heights <- rnorm(100, mean=188, sd=3)
sample(heights, size=10, replace=TRUE)
```
- `set.seed(12345) # or whatever int you want for an arg`
  - anyway : that is a way of saying "repeatably random" 
- see also : [Peng's "Simulation in R" (video)](http://www.youtube.com/watch?v=tvv4IA8PEzw)



----

Types of Data Analysis Questions
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/007typesOfQuestions/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=55) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=55_en&format=txt)

_"What kind of question are you trying to answer?"_ try categories:

- **descriptive** → describe something 
  - usually on a "census" or population 
- **exploratory** → goal : find relationships that you didn't necessarily know about
  - ok to do on a census, but more commonly on a smaller data set 
  - this type of analysis is usually where you need to remember "correlation does not equal causation"
    - (( you are looking for clues here about where to look next )) 
- **inferential** → goal : use a small sample to imply something about a larger population 
  - must: estimate the qty. you care about & the quantify the uncertainty you have about that estimate 
  - case study: look at air pollution in 500 counties across the USA, make inferences about air pollution throughout the USA
- **predictive** → goal : use the data on some objects to predict values for another object
  - REMEMBER : just because "x predicts y" does not mean that "x _causes_ y"
  - prediction is very hard 
  - case study : [Nate Silver's 538 blog](http://fivethirtyeight.blogs.nytimes.com/)
  - case study : [Forbes.com: How Target Figured Out A Teen Girl Was Pregnant Before Her Father Did](http://www.forbes.com/sites/kashmirhill/2012/02/16/how-target-figured-out-a-teen-girl-was-pregnant-before-her-father-did/)
- **causal** → goal : find out what happens to one variable when you make another variable change 
  - randomized studies more/less required 
  - usually ID'd as "average effects" 
  - "the gold standard" 
  - controlling variables & "doing experiments" so be go beyond "x _predicts_ y"
    - ( however : again : even with "causal" studies, it can be difficult to prove a causal relationship )
- **mechanistic** → goal : Understand the exact changes in variables that lead to changes in other variables for individual objects.
  - incredibly hard 
  - try to "prove" that by manipulating variable(s) that you can predict changes in other variable(s)
    - (( complex relationships w/in systems ))



----

Sources of Data Sets
--------------------------------------------------------------------------------

[slides](https://dl.dropbox.com/u/7710864/008sourcesOfDataSets/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=57) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=57_en&format=txt)

- data are defined by how they are collected:
- **census** → usually for descriptive studies 
  - census → population → measure all variables across each individual in the population 
- **observational study** → ...inferential
  - random sample from your target population
    - based on that sample, extrapolate about the population 
- **convenience sample** → all types (may be biases)
  - a sample, but not random (e.g., you can't travel too far to get your subjects)
- **randomized trial** → mandatory for causal studies 
  - get your sample, assign different treatments to different groups 
    - note: hold all other variables constant 
- (main above, other below)
- **prediction study**
  - need 2 data sets → training set (build predictive model) + test set (evaluate predictive model)
  - result of 1st set: build the predictive model
  - 2nd time around : use your predictive model and see if you can validate it 
- **studies over time** (_cross-sectional_, _longitudinal_)
  - cross-sectional: pick a time point, and measure your sample at that time point (!????)
  - longitudinal: assign individuals from your sample, and follow them over a period of time (measuring your variables again and again and again...)
- **retrospective** (_inferential_) 
  - do an "exposure" → "fast forward" to the end, and take your measurements then 
  - noted : it's inferential (not predictive, not causal)
