# [Data Analysis] Week 3 Lectures

## Example Analysis Assignment

[example file](https://d19vezwu8eufl6.cloudfront.net/dataanalysis/exampleProject.zip) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=83_en&format=txt) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=83)

"How did I do this analysis?"

* **prompt** - contains the question (naturally?)
* **assignment** - the materials that should be submitted
* **core** - has `finalcode` and `rawcode`
  * note that `raw`... contains _all_ analyses performed (not just the ones that appear in the final)
* **figures** - just the final figure
  * optionally (and/or: for yourself) consider keeping interesting exploratory figures
    * and/but - if you're using `Rmd` then you probably don't need to do this



## Exploratory Graphs (1 + 2)

slides
  [1](https://dl.dropbox.com/u/7710864/courseraPublic/week3/001exploratoryGraphs1/index.html) +
  [2](https://dl.dropbox.com/u/7710864/courseraPublic/week3/002exploratoryGraphs2/index.html) |
videos
  [1](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=85) +
  [2](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=87) |
transcripts
  [1](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=85_en&format=txt) +
  [2](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=87_en&format=txt)

* "Why do we use graphs?"
  * _communication!_
  * see patterns that you can't see w/ numbers alone
  * understand properties of the data
  * suggest modeling strategies
* characteristics of _exploratory_ graphs
  * make many of them, and quickly
  * (clean it up later)
* (( illustrating how 1 set of data could be visualized _n_ different ways... ))
* "position vs. length"
  * ask: "What is the main take-away from the graph? Is it easy to 'see what you're saying'?"
  * (is the graph easy to read?)
* "position vs. angle"
  * ("why statisticians don't like pie charts")
* whenever possible: try to use **position** as the basis for graphical comparisons

### example plots
(using the American Community Survey data from last week's quiz)

```{r}
pData <- read.csv("../quizzes/ss06pid.csv")
```

#### Boxplot
* for a quantitative variable
* goal: get an idea of the distribution of the data
* observe:
  * **median** (thick line in the middle)
  * **75th** & **25th percentile** (upper & lower bounds of the box)
  * **whiskers** - 1.5x the value of 75th & 25th percentiles, respectively
  * also: outliers appear as dots
* `varwidth` = size the width based on the number of observations (for that factor)

```{r}
boxplot(pData$AGEP, col="blue")

# "Show me AGEP but broken down by DDRS"
boxplot(pData$AGEP ~ as.factor(pData$DDRS), col="blue")

# encode additional information with `varwidth`
boxplot(pData$AGEP ~ as.factor(pData$DDRS), col=c("blue", "orange"), names=c("yes", "no"), varwidth=TRUE)
```

#### Barplot
* height of the bar = data value
  * "number of observations per class"
* continuous data?
  * break it down into chunks and look at the values that way

```{r}
barplot(table(pData$CIT), col="blue")
```

#### Histograms
* important params: breaks, freq, col, xlab, ylab, xlim, ylim, main
* "sort of like boxplots"
  * goal : quanity a univariate distribution of data
* still chunking the distribution and summing them
  * (similar to the barplot example?)
  * and/but - more fine-grained
* helps to show the shape of the distribution
  * (in that regard: more resolution on the data than w/ a boxplot)
* you can set the number of breaks

```{r}
hist(pData$AGEP, col="blue", breaks=100)
```

#### Density Plots
* like a histogram but smoothed out
* warning: could introduce errors at the boundaries
  * careful when interpretting!

```{r}
dens <- density(pData$AGEP)
plot(dens, lwd=3, col="blue")
# `lwd` = 'line width' (roughly)

# useful (vs. histogram) b/c you can overlay them
dens <- density(pData$AGEP)
densMales <- density(pData$AGEP[which(pData$SEX==1)])
plot(dens, lwd=3, col="blue")
lines(densMales, lwd=3, col="orange")
```

#### Scatterplots
* most widely used plot for exploratory analysis
* `x` & `y` are critical variables
  * (list of several other important params)
  * `?par` for more
* remember: each point represents 1 observation
* interesting: scatterplots can help illuminate weird patterns quickly
  * and/or outliers become flaringly obvious
* "size matters"
  * pay attn to scale on axis & size of dots

```{r}
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue")

# try again:
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue", cex=0.5)

# w/ color: (remember: need a number value for `col`)
plot(pData$JWMNP, pData$WAGP, pch=19, col=pData$SEX, cex=0.5)

# you can also use size to illustrate "the third variable"
percentMaxAge <- pData$AGEP/max(pData$AGEP)
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue", cex=percentMaxAge*0.5)

# get fancy: overlay lines/points
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue", cex=0.5)
lines(rep(100, dim(pData)[1]), pData$WAGP, col="grey", lwd=5)
points(seq(0, 200, length=100), seq(0, 20e5, length=100), col="red", pch=19)

# more fancy: visualizing numeric variables as factors
library(Hmisc)
ageGroups <- cut2(pData$AGEP, g=5) # <-- breaks up the age groups into "factors" (5 chunks of age groups)
plot(pData$JWMNP, pData$AGEP, pch=19, col=ageGroups, cex=0.5)
```

"If you have lots of points..." (e.g., 100k)

```{r}
# big black mass!
x <- rnorm(1e5)
y <- rnorm(1e5)
plot(x, y, pch=19)

# try sampling the values
x <- rnorm(1e5)
y <- rnorm(1e5)
sampledValues <- sample(1:1e5, size=1000, replace=FALSE)
plot(x[sampledValues], y[sampledValues], pch=19)

# ...or a "smooth scatter"
x <- rnorm(1e5)
y <- rnorm(1e5)
smoothScatter(x,y)

# ...or "hexbin"
library(hexbin)
x <- rnorm(1e5)
y <- rnorm(1e5)
hbo <- hexbin(x,y)
plot(hbo)
```

#### QQ Plots
* "Sort of like a scatter plot but with a very particular purpose..."
* plots quantiles of `x` vs. quantiles of `y`

```{r}
x <- rnorm(20)
y <- rnorm(20)
qqplot(x,y)
abline(c(0,1))
```

#### Matplot and spaghetti
* takes each column and plots it as one line
  * each column (in the matrix) becomes one specific line
* compare trends/trajectories over time

```{r}
X <- matrix(rnorm(20*5), nrow=20)
matplot(X, type="b")
```

#### Heatmaps
* "sort of like a 2D histogram"
* goal: visualize the whole matrix of data
* color --> intensity

```{r}
image(1:10, 161:236, as.matrix(pData[1:10, 161:236]))

# might want to transpose that...
newMatrix <- as.matrix(pData[1:10, 161:236])
newMatrix <- t(newMatrix)[,nrow(newMatrix):1]
image(161:236, 1:10, newMatrix)
```

#### Maps (very basics)
* very basics
  * too much to cover
* use lat/lon values w/ these

```{r}
# let's get a map
library(maps)
map("world")

# let's make up some stuff and throw it on the map
lat <- runif(40, -180, 180)
lon <- runif(40, -90, 90)
points(lat, lon, col="blue", pch=19)
```

#### Missing values and plots
* can use plots to help understand what values are missing
* gaps in the plot? check the data for `NA`!
* and/but/also : try using the `boxplot`
  * relationships b/w missing values and... values?

### Further Resources
* [R Graph Gallery](http://gallery.r-enthusiasts.com/)
* [ggplot2](http://cran.r-project.org/web/packages/ggplot2/index.html)
  * [basic introduction](http://www.r-bloggers.com/basic-introduction-to-ggplot2/)
* [lattice](http://cran.r-project.org/web/packages/lattice/index.html)
  * [introduction](http://lmdvr.r-forge.r-project.org/figures/figures.html)
* [R bloggers](http://www.r-bloggers.com/)



## Expository Graphs

[slides](https://dl.dropbox.com/u/7710864/courseraPublic/week3/003expositoryGraphs/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=89) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=89_en&format=txt)

* expository graphs are for communicating results to others
* characteristics
  * goal: communicate
  * information density = good
  * color/size are for communication (not just aesthetics)
  * needs good axes, titles, legends etc.
  * (again: COMMUNICATE CLEARLY!)

### example plots
(again: using the American Community Survey data from last week's quiz)

```{r}
pData <- read.csv("../quizzes/ss06pid.csv")
```

#### Axes
* axes need to be clear
* `xlab` + `ylab` for the axis label
* `cex.label` + `cex.axis` to adjust axis size

```{r}
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue", cex=0.5,
     xlab="Travel time (min)", ylab="Last 12 months wages ($)",
     cex.label=2, cex.axis=1.5)
```

#### Legends
* legends to clearly communicate what's on the plot
* `legend` to add the legend
* 1st 2 params: `x` + `y` (position)
  * but the values correspond to the _scales_ of the `x` + `y` axes, respectively (so: not pixels)
* reminder: _YOU_ are on the hook to make sure that the dots (etc.) in your legend match what's in your plot
  * R isn't going to do it for you

```{r}
plot(pData$JWMNP, pData$WAGP, pch=19, col="blue", cex=0.5,
     xlab="TT (min)", ylab="Wages ($)")
legend(100, 200000, legend="All surveyed", col="blue", pch=19, cex=0.5)

plot(pData$JWMNP, pData$WAGP, pch=19, cex=0.5, col=pData$SEX,
     xlab="TT (min)", ylab="Wages ($)")
legend(100, 200000, legend=c("men", "women"), col=c("black", "red"),
       pch=c(19, 19), cex=c(0.5, 0.5))
```

#### Titles
* titles: communicate the type of plot or conclusion of the plot
* the param is `main` for most plot types

```{r}
plot(pData$JWMNP, pData$WAGP, pch=19, cex=0.5, col=pData$SEX,
     xlab="CT (min)", ylab="Wages ($)",
     main="Wages earned vs. commuting time")
legend(100, 200000, legend=c("men", "women"), col=c("black", "red"),
       pch=c(19, 19), cex=c(0.5, 0.5))
```

#### Multiple panels
* if you have related plots and one to present them together
  * "to help tell the story"
* `par` is the key here
* don't overdo it! (too many becomes hard to see)

```{r}
par(mfrow=c(1,2))
hist(pData$JWMNP, xlab="CT (min)", col="blue", breaks=100, main="")
plot(pData$JWMNP, pData$WAGP, pch=19, cex=0.5, col=pData$SEX,
     xlab="CT (min)", ylab="Wages ($)")
legend(100, 200000, legend=c("men", "women"), col=c("black", "red"),
       pch=c(19, 19), cex=c(0.5, 0.5))
```

#### Adding text
* `mtext` to add some text
  * look in the function's API to see the valid values for the `side` param
* and/but : allows you to add some arbitrary text

```{r}
par(mfrow=c(1,2))
hist(pData$JWMNP, xlab="CT (min)", col="blue", breaks=100, main="")
mtext(text="(a)", side=3, line=1)
plot(pData$JWMNP, pData$WAGP, pch=19, cex=0.5, col=pData$SEX,
     xlab="CT (min)", ylab="Wages ($)")
legend(100, 200000, legend=c("men", "women"), col=c("black", "red"),
       pch=c(19, 19), cex=c(0.5, 0.5))
mtext(text="(b)", side=3, line=1)
```

#### Figure captions
* "a little bit pedantic"
* make sure to add a figure caption to explain it; something like:

> **Figure 1. Distribution of commute time and relationship to wage earned by sex (a)** and some expository text
> that goes on to explain (a) and then **(b)** we have some expository text about (b), assuming our figure has
> an (a) and a (b).

### Other considerations
What else about your audience do you need to consider before "finishing" your figures?

#### Colorblindness
* check it: [vischeck.com](http://www.vischeck.com)

### Graphical workflow
* start with rough plot (exploratory!)
* refine it (expository!)
* save it...
* `?Devices` for info

#### PDF
* `pdf(file="", heigh="", width="")`
  * `height` + `width` given in inches

```{r}
# open it:
pdf(file="twoPanel.pdf", height=4, width=8)

# business as usual (think "pipe it")
par(mfrow=c(1,2))
hist(pData$JWMNP, xlab="CT (min)", col="blue", breaks=100, main="")
mtext(text="(a)", side=3, line=1)
plot(pData$JWMNP, pData$WAGP, pch=19, cex=0.5, col=pData$SEX,
     xlab="CT (min)", ylab="Wages ($)")
legend(100, 200000, legend=c("men", "women"), col=c("black", "red"),
       pch=c(19, 19), cex=c(0.5, 0.5))
mtext(text="(b)", side=3, line=1)

# close it:
dev.off()
```

#### PNG
* almost identical to the `pdf` device commands
  * but this time: `height` + `width` are in pixels this time
* (skip the sample code this time ... it's basically the same as `pdf`)

#### dev.copy2pdf
* `dev.copy2pdf` - more/less allows you to work interactively first

```{r}
# do your regular business...
# observe your beautiful graph...

# copy exactly what you see to pdf:
dev.copy2pdf(file="twoPanelv2.pdf")
```

### Some things to avoid
* too much information
* too little
* unnecessary 3D
* poor captions

### Further resources

* [How to display data badly](http://www.jstor.org/discover/10.2307/2683253?uid=3739704&uid=2&uid=4&uid=3739256&sid=21101619120151)
* [The visual display of quantitative information](http://www.amazon.com/exec/obidos/ASIN/0961392142/7210-20)
* [Creating more effective graphs](http://www.amazon.com/exec/obidos/ASIN/047127402X/7210-20)
* [R Graphics Cookbook](http://www.amazon.com/R-Graphics-Cookbook-Winston-Chang/dp/1449316956)
* [ggplot2: Elegant Graphics for Data Analysis](http://www.amazon.com/ggplot2-Elegant-Graphics-Data-Analysis/dp/0387981403)
* [Flowing Data](http://flowingdata.com/)



## Hierarchical Clustering

[slides](https://dl.dropbox.com/u/7710864/courseraPublic/week3/004hierachicalClustering/index.html) |
[video](https://class.coursera.org/dataanalysis-001/lecture/download.mp4?lecture_id=91) |
[transcript](https://class.coursera.org/dataanalysis-001/lecture/subtitles?q=91_en&format=txt)

* exploratory analysis technique
* **clustering** in general
  * organizing things "close" into groups
  * definition of "close"?
* clustering = very important
  * used often in academics _and_ business, science, tech. etc.

### Hierarchical Clustering
* agglomerative / "bottom up"
  * find the 2 observations closest into each other
    * then merge into a single "super observation"
  * repeat!
* requires a defined distance
* requires a "merging" approach
* produces: a **dendrogram**
* **"How do we define close?"**
  * distance or similarity
    * continuous (Euclidean distance)
      * _Euclidean_ - looking (basically) for hypotenuse...
        * e.g., Baltimore lat, lon vs. D.C. lat, lon
          * figure the trigonometry out
        * √((X₁ - X₂)² + (Y₁ - Y₂)² + ... + (Z₁ - Z₂)²)
    * continuous (correlation similarity)
    * binary (Manhattan distance)
      * _Manhattan distance_ - shortest path b/w 2 points "on a grid"
      * |A₁ - A₂| + |B₁ - B₂| + ... + |Z₁ - Z₂|
  * pick a distance/similarity that makes sense w/ _your_ data

#### Example

Simple simulated data set:

```{r}
set.seed(1234); par(mar=c(0,0,0,0))
x <- rnorm(12, mean=rep(1:3, each=4), sd=0.2)
y <- rnorm(12, mean=rep(c(1,2,1), each=4), sd=0.2)
plot(x, y, col-"blue", pch=19, cex=2)
text(x+0.05, y+0.05, labels=as.character(1:12))
```

Calculate a distance b/w all of the different points:
```{r}
dataFrame <- data.frame(x=x, y=y)
dist(dataFrame)
# `dist` also takes a `method` as 2nd param ("Euclidean" is default)
```

1. "What are 2 closest points?" (points 5 & 6)
2. Merge them together.
3. Repeat! (10 & 11, merge; 9 & 12, merge; etc.)
4. Give us a dendrogram:
```{r}
dataFrame <- data.frame(x=x, y=y)
distxy <- dist(dataFrame)
hClustering <- hclust(distxy)
plot(hClustering)
```
5. Examine the dendrogram; what's close together, really?
   Where should you cut the tree? (look at the `Height` axis)

#### Example 2 ("Prettier")

```{r}
myplclust <- function( hclust, lab=hclust$labels, lab.col=rep(1, length(hclust$labels)), hang=0.1, ...) {
  # slides have copyright notice here (Eva KF Chan, 2009)
  y <- rep(hclust$height, 2)
  x <- as.numeric(hclust$merge)
  y <- y[which(x<0)]
  x <- x[which(x<0)]
  x <- abs(x)
  y <- y[order(x)]
  x <- x[order(x)]
  plot(hclust, labels=FALSE, hang=hang, ...)
  text(x=x, y=y[hclust$order]-(max(hclust$height)*hang),
       labels=lab[hclust$order], col=lab.col[hclust$order],
       srt=90, adj=c(1,0.5), xpd=NA, ...)
}
```

Above: allows you to color the leaves with a particular color...
```{r}
dataFrame <- data.frame(x=x, y=y)
distxy <- dist(dataFrame)
hClustering <- hclust(distxy)
myplclust(hClustering, lab=rep(1:3, each=4), lab.col=rep(1:3, each=4))
```

_Even_ prettier ones available... (see "R Gallery")

#### Merging
* complete
* average
* experiment : try out different ones -- which make sense?

#### Visualizing the clusters
* try `heatmap`
```{r}
dataFrame <- data.frame(x=x, y=y)
set.seed(143)
dataMatrix <- as.matric(dataFrame)[sample(1:12),]
heatmap(dataMatrix)
```

### A few more notes
* _can_ give an idea of relationships w/in the data
* picture may be unstable
  * what happens if you change a few points?
  * what happens w/ diff. missing values?
  * what happens if you pick a diff. distance?
  * what happens if you change your merge strategy?
  * what happens if you change the scale of points for one variable?
* and/but : deterministic! (functional?)
  * put same data in w/ same params: get same clustering out
* where to cut? (not always obvious)
* clustering: primarily for exploration
  * not too much explanatory power (but can lead you there)