# [Coursera] Data Analysis Assignment #1

## Prompt

> The data above consist of a sample of 2,500 peer-to-peer loans issued through
> the Lending Club (<https://www.lendingclub.com/home.action>). The interest
> rate of these loans is determined by the Lending Club on the basis of
> characteristics of the person asking for the loan such as their employment
> history, credit history, and creditworthiness scores.
>
> The purpose of your analysis is to identify and quantify associations between
> the interest rate of the loan and the other variables in the data set. In
> particular, you should consider whether any of these variables have an
> important association with interest rate after taking into account the
> applicant's FICO score. For example, if two people have the same FICO score,
> can the other variables explain a difference in interest rate between them?

[Source.](https://class.coursera.org/dataanalysis-001/human_grading/view/courses/294/assessments/4/submissions)

_Short version?_ **Given the same FICO score, can you explain the differences in interest rates?**

## Preliminaries

Load data:

```{r}
setwd('~/Desktop/coursera-data-analysis/assignments/1')
load('data/raw/loansData.rda')
head(loansData, 2)
```

~~Anything wrong with it?~~ _not yet_
```{r}
summary(loansData)
```

### What's in there?

```{r}
names(loansData)

# Qualitative data:
unique(loansData$Loan.Length)
unique(loansData$Loan.Purpose)
unique(loansData$State)
unique(loansData$Home.Ownership)
unique(loansData$FICO.Range)
unique(loansData$Inquiries.in.the.Last.6.Months)
unique(loansData$Employment.Length)

# Quantitative... so: what can we summarize with quantiles?
quantile(loansData$Amount.Requested)
boxplot(loansData$Amount.Requested)
quantile(loansData$Amount.Funded.By.Investors)
boxplot(loansData$Amount.Funded.By.Investors)

# factors?
par(las=2, mar=c(4, 4, 1, 1))
#loansData$Interest.Rate
hist(sapply(sub("%", "", loansData$Interest.Rate,), as.numeric), breaks=20)
#loansData$Debt.To.Income.Ratio
hist(sapply(sub("%", "", loansData$Debt.To.Income.Ratio,), as.numeric), breaks=35)
#loansData$Monthly.Income
hist(loansData$Monthly.Income, breaks=100)
quantile(loansData$Monthly.Income, na.rm=TRUE)
#loansData$Open.CREDIT.Lines
hist(loansData$Open.CREDIT.Lines, breaks=30)
#loansData$Revolving.CREDIT.Balance
hist(loansData$Revolving.CREDIT.Balance, breaks=50)
```

#### Ideas? Things to consider:
* Amount requested vs. amount funded?
* ~~Loan Purpose as a factor in getting funded?~~
  * _(this is tangential to the question)_
* Any loans that were _NOT_ funded?
  * `0` rows at `NA`
  * `4` rows at `0`

**What does the distribution of Loan Purpose look like?**
```{r}
par(las=2, mar=c(9, 4, 1, 1))
barplot(table(loansData$Loan.Purpose), col="blue", ylab="# loans requested")

# anything interesting here?
boxplot(loansData$Amount.Requested ~ as.factor(loansData$Loan.Purpose),
        ylab="$", varwidth=TRUE)
```
(You might be getting ahead of yourself here...)

#### Exploring the factors
```{r}
boxplot(loansData$Amount.Funded.By.Investors ~ loansData$Open.CREDIT.Lines)
boxplot(loansData$Amount.Funded.By.Investors ~ loansData$Revolving.CREDIT.Balance)
boxplot(loansData$Amount.Funded.By.Investors ~ loansData$State)
```

Going to need some buckets...

## Partition by FICO
```{r}
unique(loansData$FICO.Range)
# 43 levels ... what does it look like if collapsed to 5?
loansData.FICO.Range.floor <- cut(sapply(sub("-\\d{3}", "", loansData$FICO.Range), as.numeric), breaks=5)
table(loansData.FICO.Range.floor)
plot(loansData.FICO.Range.floor)
```

## Partition by Interest Rate
```{r}
unique(loansData$Interest.Rate)
# 430 levels

# convert nominal Interest Rate data into Ratio
loansData.Interest.Rate.as.factor <- cut(sapply(sub("%", "", loansData$Interest.Rate,), as.numeric), breaks=5)
table(loansData.Interest.Rate.as.factor)
plot(loansData.Interest.Rate.as.factor, col="blue")
```

## Exploration (1)

**Try to think about** what columns should you focus on?

Key columns to start with:
* FICO.Range
* Interest.Rate

### Any relationships between FICO range and interest range?

```{r}
par(mfrow=c(2,1))
plot(loansData.FICO.Range.floor, col="green")
plot(loansData.Interest.Rate.as.factor, col="red")
```

Looks like there might be -- similar shapes to those distributions.

As a scatter?
```{r}
par(las=2, mfrow=c(1,1), mar=c(4, 4, 1, 1))
plot(sapply(sub("-\\d{3}", "", loansData$FICO.Range), as.numeric),
     sapply(sub("%", "", loansData$Interest.Rate,), as.numeric),
     pch=19, cex=0.5, xlab="FICO range (floor)", ylab="Interest Rate (%)")
```

Generally speaking: as FICO goes up, interest rate goes down.

Does Loan Amount play a part?
```{r}
loansData$Amount.Requested.range <- cut(loansData$Amount.Requested, breaks=5)
par(las=2, mar=c(4, 4, 1, 1))
colors <- c("black", "red", "blue", "cyan", "green")
plot(sapply(sub("-\\d{3}", "", loansData$FICO.Range), as.numeric),
     sapply(sub("%", "", loansData$Interest.Rate,), as.numeric),
     col=colors,
     pch=19, cex=0.5, xlab="FICO range (floor)", ylab="Interest Rate (%)")
legend(750, 25,
       legend = unique(loansData$Amount.Requested.range),
       col = colors,
       pch=rep(19,5), cex=rep(0.5,5))
```

**Hard to tell...** With 5 levels of `Amount.Requested` it's a little hard to
see what's going on in there. Maybe with only 3 levels of `Amount.Requested`?
```{r}
loansData$Amount.Requested.range <- cut(loansData$Amount.Requested, breaks=3)
par(las=2, mar=c(4, 4, 1, 1))
colors <- c("red", "green", "blue")
plot(sapply(sub("-\\d{3}", "", loansData$FICO.Range), as.numeric),
     sapply(sub("%", "", loansData$Interest.Rate,), as.numeric),
     col=colors,
     pch=19, cex=0.5, xlab="FICO range (floor)", ylab="Interest Rate (%)")
legend(750, 25,
       legend = unique(loansData$Amount.Requested.range),
       col = colors,
       pch=rep(19,3), cex=rep(0.5,3))
```

### Try visualizing relationships of other factors to `Interest.Rate`

First: store the sanitized interest rates:
```{r}
loansData$Interest.Rate.numeric <- sapply(sub("%", "", loansData$Interest.Rate,), as.numeric)
```

#### "Amount.Requested"
```{r}
par(las=2, mar=c(4, 4, 1, 1))
plot(loansData$Interest.Rate.numeric,
     loansData$Amount.Requested,
     pch=19, cex=0.25,
     xlab="Interest Rate (%)", ylab="$ requested")
```

"Amount.Funded.By.Investors"
```{r}
par(las=2, mar=c(4, 4, 1, 1))
plot(loansData$Interest.Rate.numeric,
     loansData$Amount.Funded.By.Investors,
     pch=19, cex=0.25,
     xlab="Interest Rate (%)", ylab="$ requested")

# Any values missing?
nrow(loansData[is.na(loansData$Amount.Funded.By.Investors),])

# Any *NOT* funded?
nrow(loansData[loansData$Amount.Funded.By.Investors == 0,])

plot(loansData$Amount.Requested,
     loansData$Amount.Funded.By.Investors,
     pch=19, cex=0.25)

# what about diffs? (Requested - Funded)
loansData$Amount.diff <- apply(loansData, 1, function(row){ as.numeric(row[1]) - as.numeric(row[2])})

hist(loansData$Amount.diff)

par(mfrow=c(2,2))
hist(loansData[loansData$Amount.diff > 0,]$Amount.diff, breaks=25, main="")
hist(loansData[loansData$Amount.diff > 100,]$Amount.diff, breaks=25, main="")
hist(loansData[loansData$Amount.diff > 500,]$Amount.diff, breaks=25, main="")
hist(loansData[loansData$Amount.diff > 1000,]$Amount.diff, breaks=25, main="")
```

"Debt.To.Income.Ratio" (numeric)
```{r}
par(mfrow=c(1,1))
loansData$Debt.To.Income.Ratio.numeric <- sapply(sub("%", "", loansData$Debt.To.Income.Ratio,), as.numeric)
par(las=2, mar=c(4, 4, 1, 1))
plot(loansData$Interest.Rate.numeric,
     loansData$Debt.To.Income.Ratio.numeric,
     pch=19, cex=0.25,
     xlab="Interest Rate (%)", ylab="Debt/Income Ratio (%)")
```

"Open.CREDIT.Lines""
```{r}
# remember: loansData$Open.CREDIT.Lines comes up as a factor
par(las=2, mar=c(4, 4, 1, 1))
plot(loansData$Interest.Rate.numeric,
     loansData$Open.CREDIT.Lines,
     pch=19, cex=0.25,
     xlab="Interest Rate (%)", ylab="# open credit lines")

boxplot(loansData$Interest.Rate.numeric ~ loansData$Open.CREDIT.Lines)
```

"Revolving.CREDIT.Balance"
```{r}
# remember: loansData$Open.CREDIT.Lines comes up as a factor
par(las=2, mar=c(4, 4, 1, 1))
plot(loansData$Interest.Rate.numeric,
     loansData$Revolving.CREDIT.Balance,
     pch=19, cex=0.25,
     xlab="Interest Rate (%)", ylab="revolving credit balance ($)")
```

"Monthly.Income"
```{r}
par(las=2, mar=c(4, 4, 1, 1))
plot(loansData$Interest.Rate.numeric,
     loansData$Monthly.Income,
     pch=19, cex=0.25,
     xlab="Interest Rate (%)", ylab="monthly income ($)")

# Curious: Monthly Income vs. Open Credit Lines
plot(loansData$Open.CREDIT.Lines,
     loansData$Monthly.Income,
     pch=19, cex=0.25,
     ylab="Monthly Income ($)", xlab="# open credit lines")

# Curious: and/or :  vs. Revolving Balance
plot(loansData$Revolving.CREDIT.Balance,
     loansData$Monthly.Income,
     pch=19, cex=0.25,
     ylab="Monthly Income ($)", xlab="revolving credit bal. ($)")
```

"Inquiries.in.the.Last.6.Months"
```{r}
par(las=2, mar=c(4, 4, 1, 1))
plot(loansData$Interest.Rate.numeric,
     loansData$Inquiries.in.the.Last.6.Months,
     pch=19, cex=0.25,
     xlab="Interest Rate (%)", ylab="# credit inquiries (past 6 mo.)")

boxplot(loansData$Interest.Rate.numeric ~ loansData$Inquiries.in.the.Last.6.Months)
```

"Employment.Length"
```{r}
# 1st: fix the ordering of the factor
# http://www.r-bloggers.com/reorder-factor-levels/
loansData$Employment.Length <- factor(loansData$Employment.Length,
                                      levels(loansData$Employment.Length)[c(2:3,5:12,4,13,1)])

barplot(table(loansData$Employment.Length))
boxplot(loansData$Interest.Rate.numeric ~ loansData$Employment.Length)
```

"Home.Ownership"
```{r}
barplot(table(loansData$Home.Ownership))
boxplot(loansData$Interest.Rate.numeric ~ loansData$Home.Ownership)
```
(probably won't look any further at "Home.Ownership" as a factor)

"State"
```{r}
barplot(table(loansData$State))
boxplot(loansData$Interest.Rate.numeric ~ loansData$State)
# wow that was noisy...
```
(probably won't look any further at "State" as a factor)

"Loan.Purpose"
```{r}
par(las=2, mar=c(9, 3, 1, 1))
boxplot(loansData$Interest.Rate.numeric ~ loansData$Loan.Purpose)
# might be something here; look at:
# - car, educational, renewable_energy
```

(won't waste time with "Loan.Length")

### "Uh... what else?"

Heatmap?
```{r}
set.seed(282147)
hmSample <- loansData[sample(nrow(loansData), 100),]
heatmap(matrix(c(hmSample$Amount.Requested, hmSample$Amount.Funded.By.Investors, hmSample$Interest.Rate.numeric, hmSample$Debt.To.Income.Ratio.numeric, hmSample$Monthly.Income), nrow=100, ncol=5))
```