# [Coursera] Data Analysis Assignment #1

## Prompt

> The data above consist of a sample of 2,500 peer-to-peer loans issued through
> the Lending Club (<https://www.lendingclub.com/home.action>). The interest
> rate of these loans is determined by the Lending Club on the basis of
> characteristics of the person asking for the loan such as their employment
> history, credit history, and creditworthiness scores.
>
> The purpose of your analysis is to identify and quantify associations between
> the interest rate of the loan and the other variables in the data set. In
> particular, you should consider whether any of these variables have an
> important association with interest rate after taking into account the
> applicant's FICO score. For example, if two people have the same FICO score,
> can the other variables explain a difference in interest rate between them?

[Source.](https://class.coursera.org/dataanalysis-001/human_grading/view/courses/294/assessments/4/submissions)

_Short version?_ **Given the same FICO score, can you explain the differences in interest rates?**

## Load Data
```{r}
setwd('~/Desktop/coursera-data-analysis/assignments/1')
load('data/raw/loansData.rda')
```

## Clean Data
```{r}
# `%` to numeric
loansData$Interest.Rate <- sapply(sub("%", "", loansData$Interest.Rate,), as.numeric)
loansData$Debt.To.Income.Ratio <- sapply(sub("%", "", loansData$Debt.To.Income.Ratio,), as.numeric)

# Loan.Length
loansData$Loan.Length <- sapply(sub(" months", "", loansData$Loan.Length,), as.numeric)

# fix Employment.Length factor
loansData$Employment.Length <- factor(loansData$Employment.Length,
                                      levels(loansData$Employment.Length)[c(2:3,5:12,4,13,1)])

# coerce Employment.Length
# "< 1" becomes 0; "10+" becomes 10; "n/a" becomes -1
loansData$Employment.Length.numeric <- sub(" years?", "", loansData$Employment.Length)
loansData$Employment.Length.numeric <- sub("\\+", "", loansData$Employment.Length.numeric)
loansData$Employment.Length.numeric <- sub("< 1", "0", loansData$Employment.Length.numeric)
loansData$Employment.Length.numeric <- sub("n/a", "-1", loansData$Employment.Length.numeric)
loansData$Employment.Length.numeric <- sapply(loansData$Employment.Length.numeric, as.numeric)

# coerce FICO range into something quantitative:
loansData$FICO.Range.floor <- sapply(sub("-\\d{3}", "", loansData$FICO.Range), as.numeric)
```

## SVD

With or without the qualitative factors? Prep it first:

```{r}
loansData.for.SVD <- loansData

# drop the columns we already coerced
drops <- c("FICO.Range", "Employment.Length")
loansData.for.SVD <- loansData.for.SVD[,!(names(loansData.for.SVD) %in% drops)]

# convert factors to numerics
converts <- c("Loan.Purpose", "State", "Home.Ownership")
for (n in converts) {
  loansData.for.SVD[[n]] <- as.numeric(loansData.for.SVD[[n]])
}

# get rid of `NA`
for (n in names(loansData.for.SVD)) {
  loansData.for.SVD <- loansData.for.SVD[!is.na(loansData.for.SVD[[n]]),]
}

# drop FICO (since we're holding it constant)
# drop Interest.Rate (since that is effectively our dependent variable)
drops <- c("FICO.Range.floor", "Interest.Rate")
loansData.for.SVD <- loansData.for.SVD[,!(names(loansData.for.SVD) %in% drops)]
```

First:

### With the Qualitative Factors

```{r}
svd1 <- svd(scale(loansData.for.SVD))
par(mfrow=c(1,2))
plot(svd1$d,xlab="Column",ylab="Singular value",pch=19)

## the crux: (get the % variance explained)
plot(svd1$d^2/sum(svd1$d^2),xlab="Column",ylab="Percent of variance explained",pch=19)
```

What's above 10%?

* "Amount.Requested"
* "Amount.Funded.By.Investors"
* "Loan.Length"

### _Without_ the Qualitative Factors
```{r}
# since these are meaningless when converted to numerics
drops <- c("Loan.Purpose","State","Home.Ownership")
loansData.for.SVD <- loansData.for.SVD[,!(names(loansData.for.SVD) %in% drops)]

svd2 <- svd(scale(loansData.for.SVD))
par(mfrow=c(1,2))
plot(svd2$d,xlab="Column",ylab="Singular value",pch=19)

## the crux: (get the % variance explained)
plot(svd2$d^2/sum(svd2$d^2),xlab="Column",ylab="Percent of variance explained",pch=19)
```

What's above 10%?

* "Amount.Requested"
* "Amount.Funded.By.Investors"
* "Loan.Length"
* "Debt.To.Income.Ratio"
* "Monthly.Income"

(So not a whole lot of surprises!)

## Baseline: FICO by Interest.Rate

```{r}
par(mfrow=c(1,1))
# we already know: as FICO goes down, Interest goes up:
smoothScatter(loansData$FICO.Range.floor, loansData$Interest.Rate)
```

Telling us what we already know: lower FICO, higher interest rate.

## Loan.Length?

FICO by Interest.Rate, colored by Loan.Length

```{r}
par(mfrow=c(1,1))
plot(loansData$FICO.Range.floor,
     loansData$Interest.Rate,
     # coerce months to years (so they aren't the same color)
     col=sapply(loansData$Loan.Length, function(x) { x / 12 }),
     pch=19, cex=0.5)
```

Longer loans trend toward higher rates.

## Debt.To.Income.Ratio?

FICO by Interest.Rate, colored by Debt.To.Income.Ratio

```{r}
par(mfrow=c(1,1))
plot(loansData$FICO.Range.floor,
     loansData$Interest.Rate,
     col=cut(loansData$Debt.To.Income.Ratio, breaks=3),
     pch=19, cex=0.5)
```

## "Monthly.Income"?

FICO by Interest.Rate, colored by Monthly.Income

```{r}
par(mfrow=c(1,1))
plot(loansData$FICO.Range.floor,
     loansData$Interest.Rate,
     # the number of breaks doesn't seem to make much difference
     # ...not in how the colors turn out... 
     col=cut(loansData$Monthly.Income, breaks=5),
     pch=19, cex=0.5)
```

## Linear Model?
```{r}
lmNoAdjust <- lm(loansData$FICO.Range.floor ~ loansData$Interest.Rate)

plot(loansData$Interest.Rate, lmNoAdjust$residuals,
     col=sapply(loansData$Loan.Length, function(x) { x / 12 }),
     pch=19, cex=0.5)
```
