# [Coursera] Data Analysis Assignment #1

## Prompt

> The data above consist of a sample of 2,500 peer-to-peer loans issued through
> the Lending Club (<https://www.lendingclub.com/home.action>). The interest
> rate of these loans is determined by the Lending Club on the basis of
> characteristics of the person asking for the loan such as their employment
> history, credit history, and creditworthiness scores.
>
> The purpose of your analysis is to identify and quantify associations between
> the interest rate of the loan and the other variables in the data set. In
> particular, you should consider whether any of these variables have an
> important association with interest rate after taking into account the
> applicant's FICO score. For example, if two people have the same FICO score,
> can the other variables explain a difference in interest rate between them?

[Source.](https://class.coursera.org/dataanalysis-001/human_grading/view/courses/294/assessments/4/submissions)

## Preliminaries

Load data:

```{r}
setwd('~/Desktop/coursera-data-analysis/assignments/1')
load('data/raw/loansData.rda')
head(loansData)
```

~~Anything wrong with it?~~ _not yet_
```{r}
summary(loansData)
```

What's in there?
```{r}
names(loansData)
unique(loansData$Amount.Requested)
unique(loansData$Amount.Funded.By.Investors)
unique(loansData$Interest.Rate)
unique(loansData$Loan.Length)
unique(loansData$Loan.Purpose)
unique(loansData$Debt.To.Income.Ratio)
unique(loansData$State)
unique(loansData$Home.Ownership)
unique(loansData$Monthly.Income)
unique(loansData$FICO.Range)
unique(loansData$Open.CREDIT.Lines)
unique(loansData$Revolving.CREDIT.Balance)
unique(loansData$Inquiries.in.the.Last.6.Months)
unique(loansData$Employment.Length)
```